\Title

### 2013-08-05 synmut.tex
added a few references and discussions of issues ref 2 raised. worked on response

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Computer simulations}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Computer simulations were performed using FFPopSim
\citep{zanini_ffpopsim:_2012}. Briefly, FFPopSim enables individual-based
simulations where each site in the genome is represented by one bit that can be
in one of two states. Outcrossing rates, crossover rates, mutations rates and
arbitrary fitness functions can be specified. We used an outcrossing rate of
$r=0.01$ per day \citep{batorsky_estimate_2011, neher_recombination_2010},
a mutation rate of $\mu=2 \times 10^{-5}$ \citep{mansky_lower_1995, abram_nature_2010}.
We performed the simulations with
generation times of both 1 and 2 days, because this parameter varies across patients
\citep{perelson_hiv-1_1996, perelson_decay_1997}; our conclusions are shown in
did not change. We simulated intrapatient
evolution for 6000 days. For simplicity, third positions of every codon were
deemed synonymous and assigned either a selection coefficient $0$ with
probability $1-\alpha$ or a deleterious effect $s_d$ with probability $\alpha$.
Mutations at the first and second positions were assigned strongly deleterious 
fitness effects 0.02. At 
rate $k_A$, a random locus in the genome is designated an epitope that can
escape by one or several mutations with exponentially distributed escape rates
with mean $\epsilon$. Both full-length HIV-1 genomes and \env{}-only simulations
were performed and yielded comparable results.

The simulations were repeated 1000 times with a four dimensional Latin hypercube sampling scheme.
For two parameters, ranges were chosen around estimates from the recent literature:
the average escape rate $\epsilon$ of escape mutation was sampled such that its logarithm was
uniform between $2 \cdot 10^{-2.5}$ and $2 \cdot 10^{-1.5}$
\citep{neher_recombination_2010} and the rate $k_A$ of new antibody
challenges such that its logarithm was uniform between $10^{-3}$ and $10^{-2}$
per generation \citep{richman_rapid_2003, moore_limited_2009}.
For the other two parameters, ranges were determined to include all simulations
that reproduced the experimental data: the fraction of deleterious sites $\alpha$ was sampled uniformly
between 0.75 and 1.0; the average deleterious effect $s_d$ was sampled such that
its logarithm was uniformly distributed  between $2 \cdot 10^{-4}$ and $2 \cdot
10^{-2}$. Populations were initialized with a homogeneous founder
population and were kept at an average size of $N=10^4$ throughout the
simulation. After 30 generations of burn-in to create genetic diversity, new
epitopes were introduced at a constant rate $k_A$.

@@

For each set of parameters, fixation probabilities and probabilities of
synonymous polymorphisms $P_\text{interm}$ were calculated as averages over
100 repetitions (with different random seeds).

@@

If the immune system
recognizes the escape mutant before its fixation, the mutant might cease to be
beneficial and disappear soon, despite its quick initial rise in frequency. In
support of this idea, in \citet{richman_rapid_2003, bunnik_autologous_2008},
antibody responses against escape mutants have been reported. These responses are
delayed by a few months, roughly matching the average time needed by an escape
mutant to rise from low to high frequency. To model this type of behavior, we
assumed that antibody responses against escape SNVs arise with a rate
proportional to the frequency of the escape variant and abolish the benefit of the
escape mutations. As expected, this type of time-dependent selection retained the
potential for hitchhiking, but reduced fixation of nonsynonymous SNVs.
\figurename~S\timedependence~shows that $\pfix$ of synonymous SNVs is not
affected by this change, while $\pfix$ of nonsynonymous SNVs approaches the
diagonal as the rate of recognition of escape mutants is increased.

Several different escape mutations that arise within the 
same epitope have similar effects. Their benefits
are not additive, because each of them is essentially sufficient to escape and
no additional benefit is gained from combining them. As a consequence, several
escape SNVs rise to high frequency rapidly, while the one with the smallest cost
in terms of replication, packaging, etc.~is most likely to eventually fix, while
all others are lost. The emergence of multiple competing escape variants in HIV-1
infections has been shown \citep{moore_limited_2009, bar_early_2012}. 
Similarly, this scenario has been explicitly observed in the evolution of
resistance to 3TC, where the mutation M184V is often preceded by M184I
\citep{hedskog_dynamics_2010}. We implemented within epitope competition 
in the model by allowing for multiple escape mutations per epitope that do
not provide additional benefit to the virus when combined. Again, we found that the potential for
hitchhiking is retained by within-epitope competition but that the
fixation probability of nonsynonymous SNVs is reduced. With roughly six
mutations per epitope, the simulation data are compatible with observations; see
\figurename~S\withinepi. The two scenarios, time-dependent selection and
competition between equivalent escape pathways, are not exclusive and possibly
both important in HIV-1 evolution.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Discussion}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
By analyzing the fate of single nucleotide variants (SNVs) in
longitudinal data of HIV-1 \env{} evolution, we demonstrated selection
against synonymous substitutions in the relatively conserved regions
C2-C4. We suggest through computational modeling that these
SNVs have deleterious effects on the order of $0.003$ and that they are
brought to high frequency through linkage to adaptive mutations.
Comparison with biochemical data (SHAPE) of base pairing propensity in the RNA
genome of HIV-1 indicated that these mutations tend to disrupt RNA secondary structures\citep{watts_architecture_2009}. Computer models
of RNA folding predict stable hairpins in these regions that
have been suggested to be functional and termed ``insulating
stems'' \citep{watts_architecture_2009, sanjuan_interplay_2011}.
The weak selection against synonymous variants is compatible with the
negative results of \textit{In vitro} replication assays investigating
the fitness effects of small RNA hairpins in HIV-1
\citep{knoepfel_role_2013}: it would take hundreds of cell culture
passages to detect fitness effects of the order of one per mille. The longitudinal data,
however, spans many years, hence our analysis is able to quantify the
subtle fitness effect of RNA structure within single infections.

The fixation probabilities and the sojourn times of SNVs represent a rich and simple
summary statistics: they are useful to characterize longitudinal sequence data and
can be compared to models via computer simulations. These statistics are informative
even in absence of a neutral control and thus appropriate to analyze
properties of synonymous sites. 
We find selection against synonymous substitutions despite the
fact that the corresponding sites are not strongly conserved in
cross sectional data. This is consistent with a recent comparative
analysis of SIV and HIV-1 RNA secondary structure using SHAPE assays and
computational methods \citep{pollom_comparison_2013}. While large scale
patterns of RNA structures tend to agree in both viruses, the individual
base pairs forming the structures are almost always
discordant. Even though the molecular architecture of these structures
changes over time, selection seems to maintain them, which reduces the
fixation probability and hence the rate of evolution at synonymous
sites. As expected from this argument, the evolutionary rate
at synonymous sites varies greatly along the HIV-1 genome
\citep{mayrose_towards_2007} (see also \figurename~S\syndiv). This variation can confound estimates of
selection on proteins substantially \citep{ngandu_extensive_2008}. 
The dynamic nature of HIV secondary structure makes it hard to find the 
compensatory mutations that would restore base pairing in the
longitudinal data; in fact, the exact base pairing pattern is
most likely different than in the reference sequence. Nevertheless, the
importance of secondary structure underscores the realization that viral
fitness is a complicated function (fitness landscape) in the
high-dimensional space of possible genotypes
\citep{ferguson_translating_2013}. We show, that this landscape is shaped not
only by interactions between amino-acids, but also by ubiquitous
interactions between nucleotides in RNA structures.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Computer simulations}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Computer simulations were performed using FFPopSim
\citep{zanini_ffpopsim:_2012}. Briefly, FFPopSim enables individual-based
simulations where each site in the genome is represented by one bit that can be
in one of two states. Outcrossing rates, crossover rates, mutations rates and
arbitrary fitness functions can be specified. While the best estimates
for the HIV generation time is roughly two days
\citep{perelson_hiv-1_1996, markowitz_novel_2003}, the generation time
has very little influence on the result and basically sets the unit of
time in which other parameters are measured. For simplicity, we use one
generation per day and have checked that the results obtained are
indistinguishable from simulations with a generation time of two days.

\subsubsection*{Parameters}
In order to simulate HIV, we have to specify several parameters that are
known to various degrees. The mutation rate has been measured using
exogenous LacZ constructs and the average nucleotide substitution rate is 
estimated to be $\mu\approx 2 \times 10^{-5}$ per generation \citep{mansky_lower_1995,
abram_nature_2010}. In our simulations, we vary the mutations rate
from $?? -?? \times 10^{-5}$ per day. Different virions within an
infected person can recombine their genome by coinfection of the same
cell, followed by template switching during reverse transcription. 
Template switching happens between around ten times per reverse
transcription \citep{levy_dynamics_2004}. The combined recombination rate of template
switching and coinfection has been estimated based on modeling to be around $\rho\approx
10^{-5}$ per nucleotide and day
\citep{neher_recombination_2010,batorsky_estimate_2011}, implying a
coinfection rate of a few percent or less, as has been recently
confirmed experimentally \citep{josefsson_single_2013}. We assume a
template switching rate of $10^{3}$ per nucleotide and vary the
recombination rate between $??-??\times 10^{-5}$ by adjusting the
coinfection rate. The population size relevant to evolution in chronic
infection has been hotly debated. The number of cells infected by HIV-1
per day is on the order of $10^{7}$ \citep{perelson_hiv-1_1996}, but the
number of viruses contributing to the next generation might be
considerably smaller due to the burstiness of replication. 
Recent evidence points to a relevant population
size in excess of the inverse mutation rate
\citep{boltz_ultrasensitive_2012}. In fact, the evolutionary dynamics
depends only weakly on the actual value of the population size once the
product $N\mu$ is of order one or larger \citep{neher_genetic_2011}. We
simulate populations between $N=10^4$ and $10^5$. Larger populations
become prohibitively costly to simulate.

Another crucial ingredient for our simulations is the fitness landscape,
i.e., the effects of mutations on fitness and possible interactions
between mutations. For simplicity, third positions of every codon were
deemed synonymous and assigned either a selection coefficient $0$ with
probability $1-\alpha$ or a deleterious effect $s_d$ with probability $\alpha$.
Mutations at the first and second positions were assigned strongly deleterious 
fitness effects ??? COULD MAKE THIS STRONGER, NEED CITATION. At 
rate $k_A$, a random locus in the genome is designated an epitope that can
escape by one or several mutations with exponentially distributed escape rates
with mean $\epsilon$. The rate of escape in chronic infection is
reported to be on the order of a few percent
\citep{Asquith:2006p28003,moore_limited_2009}, consistent with the
finding that virus is neutralized by serum from 3-6 month later
\citep{richman_rapid_2003}. We simulated escape rates between $2 \cdot
10^{-2.5}$ and $2 \cdot 10^{-1.5}$ per day. 
The rate at which new anti-body challenges arise is more difficult to
quantify, but the majority of non-synonymous substitutions in gp120 are
probably driven by escape. Since the sequence of the C2-V5 region of
gp120  changes at a rate of about 1\% per year
\citep{shankarappa_consistent_1999}, this rate should be on the order of
$k_A=0.003$ per day.

@@

\subsubsection*{Sampling and analysis}
To obtain reasonable sampling of a particular parameter combination we
ran simulation for 6000 days and we repeated each run 100 times with
different seeds for the random number generator. Both full-length HIV-1
genomes and \env{}-only simulations were performed and yielded
comparable results.  Populations were initialized with a homogeneous
founder population and were kept at an average size of $N=10^4$
throughout the simulation. After 30 generations of burn-in to create
genetic diversity, new epitopes were introduced at a constant rate
$k_A$.  The simulations were repeated 1000 times with a four dimensional
Latin hypercube sampling scheme \citep{mckay_comparison_1979}.  For each
parameter, 1000 random values were chosen from a distribution uniform in
log-space and bounded by the ranges given for each of these parameters
above. These 1000 values for each parameter were combined using a Latin
hypercube sampling scheme.

@@

If the immune system recognizes the escape mutant before its fixation,
the mutant might cease to be beneficial and disappear soon, despite its
quick initial rise in frequency. In support of this idea, in
\citet{richman_rapid_2003,wei_antibody_2003,bunnik_autologous_2008},
antibody responses against escape mutants have been reported. These
responses are delayed by a few months, roughly matching the average time
needed by an escape mutant to rise from low to high
frequency. Furthermore, neutralization responses against early virus
sometimes decline later in infection, which could lead to a reversion of
some sites. To model
this type of behavior, we assumed that antibody responses against escape
SNVs arise with a rate proportional to the frequency of the escape
variant and abolish the benefit of the escape mutations. As expected,
this type of time-dependent selection retained the potential for
hitchhiking, but reduced fixation of nonsynonymous SNVs.
\figurename~S\timedependence~shows that $\pfix$ of synonymous SNVs is
not affected by this change, while $\pfix$ of nonsynonymous SNVs
approaches the diagonal as the rate of recognition of escape mutants is
increased.

Several different escape mutations that arise within the same epitope
have similar effects. Their benefits are not additive, because each of
them is essentially sufficient to escape and no additional benefit is
gained from combining them. As a consequence, several escape SNVs rise
to high frequency rapidly, while the one with the smallest cost in terms
of replication, packaging, etc.~is most likely to eventually fix, while
all others are lost. The emergence of multiple competing escape variants
in HIV-1 infections has been shown \citep{moore_limited_2009,
  bar_early_2012,wei_antibody_2003}.  This scenario has been explicitly
observed in the evolution of resistance to 3TC, where the mutation M184V
is often preceded by M184I \citep{hedskog_dynamics_2010} or in the
context of CTL escape mutations where multiple variants rise to high
frequencies rapidly but most of the are lost again
\citep{fischer_transmission_2010}. We implemented within epitope
competition in the model by allowing for multiple escape mutations per
epitope that do not provide additional benefit to the virus when
combined. Again, we found that the potential for hitchhiking is retained
by within-epitope competition but that the fixation probability of
nonsynonymous SNVs is reduced. With roughly six mutations per epitope,
the simulation data are compatible with observations; see
\figurename~S\withinepi. The two scenarios, time-dependent selection and
competition between equivalent escape pathways, are not exclusive and
possibly both important in HIV-1 evolution.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Discussion}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
By analyzing the fate of single nucleotide variants (SNVs) in
longitudinal data of HIV-1 \env{} evolution, we demonstrated selection
against synonymous substitutions in the relatively conserved regions
C2-C4. We suggest through computational modeling that these
SNVs have deleterious effects on the order of $0.003$ and that they are
brought to high frequency through linkage to adaptive mutations.
While both the rise and fall of synonymous SNV is subject to hichhiking,
the consistent loss of these variants conditional on being at high
frequency implies selection against these SNVs.
Comparison with biochemical data (SHAPE) of base pairing propensity in the RNA
genome of HIV-1 indicated that these mutations tend to disrupt RNA secondary structures\citep{watts_architecture_2009}. Computer models
of RNA folding predict stable hairpins in these regions that
have been suggested to be functional and termed ``insulating
stems'' \citep{watts_architecture_2009, sanjuan_interplay_2011}.
The weak selection against synonymous variants is compatible with the
negative results of \textit{In vitro} replication assays investigating
the fitness effects of small RNA hairpins in HIV-1
\citep{knoepfel_role_2013}: it would take hundreds of cell culture
passages to detect fitness effects of the order of one per mille. The longitudinal data,
however, spans many years, hence our analysis is able to quantify the
subtle fitness effect of RNA structure within single infections. 
The fixation probabilities and the sojourn times of SNVs represent a
rich and simple summary statistics of longitudinal sequence data. Most
importantly, these statistics are informative even in absence of a
neutral control and thus appropriate to analyze properties of synonymous
sites.

We find selection against synonymous substitutions despite the
fact that the corresponding sites are not strongly conserved in
cross sectional data. This is consistent with a recent comparative
analysis of SIV and HIV-1 RNA secondary structure using SHAPE assays and
computational methods \citep{pollom_comparison_2013}. While large scale
patterns of RNA structures tend to agree in both viruses, the individual
base pairs forming the structures are almost always
discordant. Even though the molecular architecture of these structures
changes over time, selection seems to maintain them, which reduces the
fixation probability and hence the rate of evolution at synonymous
sites. As expected from this argument, the evolutionary rate
at synonymous sites varies greatly along the HIV-1 genome
\citep{mayrose_towards_2007} (see also \figurename~S\syndiv). This variation can confound estimates of
selection on proteins substantially \citep{ngandu_extensive_2008}. 
The dynamic nature of HIV secondary structure makes it hard to find the 
compensatory mutations that would restore base pairing in the
longitudinal data; in fact, the exact base pairing pattern is
most likely different than in the reference sequence. Nevertheless, the
importance of secondary structure underscores the realization that viral
fitness is a complicated function (fitness landscape) in the
high-dimensional space of possible genotypes
\citep{ferguson_translating_2013}. We show, that this landscape is shaped not
only by interactions between amino-acids, but also by ubiquitous
interactions between nucleotides in RNA structures.
